### 2017-02-14 2017mergingmerges/section/01-Introduction.tex
mainly prose

The above example seems innocuous, but existing fusion transformations cannot handle it.
As observed by \citet{kay2009you}, both pull- and push-based fusion have fundamental limitations.
Pull-based systems such as stream fusion~\cite{coutts2007stream} cannot perform fusion when a particular array or intermediate result is used by multiple consumers.
We refer to this situation as a \emph{split} --- when the example above is viewed as a data-flow network the flow from the argument array @arr1@ is split into both the @group@ and @merge@ consumers.
Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but in this case splits can cause excessive inlining which duplicates work.
Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse this example because they do not support combinators with multiple inputs.
We refer to such a situation as a \emph{join} --- in the example above the @merge@ combinator expresses a join in the data-flow graph.
Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of combinators.
More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse this case, but requires the program to be rewritten to use explicitly polarized stream types, rather than the more familiar arrays.
Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent combinators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.

@@

\ben{Add a few more sentences on related work. Explain how this work extends the old flow fusion paper. It is not short-cut fusion like Oleg's recent work. We are not in the same space as Fortran style array fusion transformations like polyhedral}
\amos{I'm not convinced anybody reading this cares about Fortran or polyhedral}
\ben{I agree. I meant that we should cite related work to say that's also called ``fusion'', but in a completely different domain.}

>>>>

The above example seems innocuous, but existing fusion transformations cannot handle it.
As observed by \citet{kay2009you}, both pull- and push-based fusion have fundamental limitations.
Pull-based systems such as stream fusion~\cite{coutts2007stream} cannot perform fusion when a particular array or intermediate result is used by multiple consumers.
We refer to this situation as a \emph{split} --- when the example above is viewed as a data-flow network the flow from the argument array @arr1@ is split into both the @group@ and @merge@ consumers.
Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but in this case splits can cause excessive inlining which duplicates work.
Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse this example because they do not support combinators with multiple inputs.
We refer to such a situation as a \emph{join} --- in the example above the @merge@ combinator expresses a join in the data-flow graph.
Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of combinators.
More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse this case, but requires the program to be rewritten to use explicitly polarized stream types, rather than the more familiar arrays.
Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent combinators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
The polyhedral model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level and, other than conservative extensions, only supports affine loops, which rules out combinators such as @group@ and @merge@.

@@

\ben{Add a few more sentences on related work. Explain how this work extends the old flow fusion paper. It is not short-cut fusion like Oleg's recent work. We are not in the same space as Fortran style array fusion transformations like polyhedral}

### 2017-02-14 2017mergingmerges/section/XX-Processes.tex
mainly prose

Each combinator defines a stream process which is expressed in terms of a simple imperative program with a local heap. The process pulls data from an aribrary number if input streams and pushes data to at least one output stream. \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.

@@

\begin{center}
\begin{alltt}
           group 
             = \(\lambda\) (s1: Stream a) (s2: Stream a). 
               \(\nu\) (first: Bool)  (last: a) (value: a) (A0..A3: Label).
\end{alltt}
\begin{code}
               process
               { ins:    { s1 }
               , outs:   { s2 }
               , heap:   { f = True, l = 0, v = 0 }
               , label:  A0
               , instrs: { A0 = pull s1 v            A1 {}
                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
                         , A2 = push s2 v            A3 { l = v, f = F }
                         , A3 = drop s1              A0 {} } }
\end{code}
\end{center}
\vspace{1em}
\includegraphics[scale=1.1]{figures/state-group.pdf}
\caption{The group combinator}
\label{fig:Process:Group}
\end{figure}

@@

The @group@ combinator has two parameters, @s1@ and @s2@ which bind the input and output streams respectively. The \emph{nu-binders} like \mbox{$\nu$ @(first: Bool)@...} indicate that each time we instantiate the @group@ combinator we should create fresh names for @first@, @last@ and so on that do not conflict with other instantiations.

The body of the combinator is a record that defines the process. The @ins@ field of the record defines the set of input streams and the @outs@ field the set of output streams. The @heap@ field gives the starting state of each of the local variables. The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the first instruction.

The instruction @(pull s1 value A1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @v@ (value), then continues with the instruction at label @A1@. In the syntax the set @{}@ at the end of the instruction can be used to update values in the heap, but as we do not need to perform heap updates in this instance we leave it empty.

The instruction @(case (f || (l /= v)) L2 {} L3 {})@ checks whether the predicate @(f || l /= v)@ is true and if so continues with the instruction at label @A2@, otherwise continues with the instruction at label @A3@. We use variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.

@@

Finally, the instruction @(drop s1 A0 {})@ signals that the current element that we pulled from stream @s1@ is no longer required, and contines with the instruction at @A0@. This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm but has no direct operational meaning. We discuss @drop@ further in \TODO{ref}.

@@

As this process merges infinite streams, if we execute it using a finite prefix then the final state will be an intermediate one that may not yet have pushed all available output. For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output. The machine will arrive at instruction @L?@ which blocks waiting for the next value to be pulled from @arr1@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}

@@

% -----------------------------------------------------------------------------
\clearpage{}
\subsection{Fusion}
Our fusion algorithm takes two process state machines and produces a new one that performs the job of both. For example, suppose we need to produce a single machine that performs the job of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}. We want a result machine that performs the job of a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@. Relative to the @uniquesUnion@ example we will associate a stream named @s1@ with the first array (@arr2@) and a stream named @s2@ with the second.

@@

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'. The `none` state means that we have not yet pulled a value from the associated stream. The `pending` state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have` state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine. In Fig.\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1` is set to `have` in @F3@ after we have set `x1 = b1'.

@@

% -----------------------------------------------------------------------------
\subsection{Fused Result}
We are now ready to take our @group@ and @merge@ processes and fuse them together. 
We create a new process with inputs @file1@ and @file2@ and outputs @uniques@ and @merged@.
In order to make the fused process easier to understand, we have performed some minor optimisations on the output of the fusion algorithm described in~\S\ref{s:Fusion}, such as removing @jump@s and merging the two variables @value@ and @a@ into a single variable @value_a@.
Rather than explaining the process in detail, we simply note that the fused process is an interleaved scheduling of the two, starting with executing @group@, then executing @merge@, and going back to @group@ whenever pulling from @file1@ again.
The labels @L0@-@L2@ and @A1@-@A4@ are much the same as those in @group@, while @C0@, @A0@-@A2@ and @B0@-@B2@ are similar to those in @merge@.

\begin{code}
process (group/merge)
     ins: file1 file2
    outs: uniques merged
    heap: {first = True, last, value_a, b}
   label: L0
  blocks: L0 = pull file1   value_a               L1
          L1 = case (or first (last /= value_a))  L2    L3
          L2 = push uniques value_a               L3{last = value_a, first = False}
          L3 = pull file2 b                       C0

          C0 = case (value_a < b)                 A0    B0

          A0 = push merged  value_a               A1
          A1 = drop file1                         A2
          A2 = pull file1   value_a               A3
          A3 = case (or first (last /= value_a))  A4    C0
          A4 = push uniques value_a               C0{last = value_a, first=False}

          B0 = push merged b                      B1
          B1 = drop file2                         B2
          B2 = pull file2  b                      C0
\end{code}

@@

A process defines a stream computation, taking any number of input streams and producing at least one output stream.
Processes have an optional name in parentheses used only for descriptive purposes, and does not need to be unique.
The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
The input states are explained in detail later in \S\ref{s:Process:Eval}.

The output streams are in some sense ``owned'' by the process that produces them.
While a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
A process may, however, produce multiple output streams.

The heap is used for evaluation of expressions.
Each process has its own private heap, meaning that the only communication between processes occurs by streams.

The label is the current block, and each block is the instruction to be executed in the current state.
Instructions can pull from a stream, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.

@@

All instructions take as an argument the next label state to jump to, which includes any variable updates that should be performed on the private heap at the same time.
Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.

A process network is a set of multiple processes that can be evaluated concurrently.
The intersection of all process outputs should be empty - there should be no overlap.
Any inputs that are not mentioned as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

@@

We now describe evaluation of processes and process networks.
We split evaluation into three main parts:
\begin{itemize}
\item Injection, where values are inserted into a process' input buffer.
Injection is only possible when the process input buffer is empty.
\item Shaking, where a process takes a step from one label to another.
Shaking a process results in a new process as well as an output message.
If a process pushes to a stream, the push value must be able to be injected to other processes.
\item Feeding, where an environment of input values are fed to the processes, and output values are collected.
This is the `top level' of evaluation that uses both injection and shaking.
\end{itemize}

>>>>

Each combinator defines a stream process which is expressed in terms of a simple imperative program with a local heap.
The process pulls data from an arbitrary number of input streams and pushes data to at least one output stream.
Each array is converted to its own process with a single output stream, which repeatedly pushes the values from the array.
% \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.

% make it explicit that the processes are only an intermediate repr
The process language is used for the implementation of the combinators and as an intermediate representation for fusion, while the orignal program using these combinators can be written without any knowledge of processes.

@@

\begin{center}
\begin{alltt}
           group 
             = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat). 
               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
\end{alltt}
\begin{code}
               process
               { ins:    { s1 }
               , outs:   { s2 }
               , heap:   { f = T, l = 0, v = 0 }
               , label:  A0
               , instrs: { A0 = pull s1 v            A1 {}
                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
                         , A2 = push s2 v            A3 { l = v, f = F }
                         , A3 = drop s1              A0 {} } }
\end{code}
\end{center}
\vspace{1em}
\includegraphics[scale=1.1]{figures/state-group.pdf}
\caption{The group combinator}
\label{fig:Process:Group}
\end{figure}

@@

The @group@ combinator has two parameters, @s1@ and @s2@, which bind the input and output streams respectively. The \emph{nu-binders} \mbox{$\nu$ @(f: Bool) (l: Nat)@...} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations.

The body of the combinator is a record that defines the process. The @ins@ field of the record defines the set of input streams and the @outs@ field the set of output streams. The @heap@ field gives the initial values of each of the local variables. The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the first instruction.

The instruction @(pull s1 v A1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @v@ (value), then continues with the instruction at label @A1@. In the syntax the set @{}@ at the end of the instruction can be used to update values in the heap, but as we do not need to perform heap updates in this instance we leave it empty.

The instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || l /= v)@ is true and if so continues with the instruction at label @A2@, otherwise continues with the instruction at label @A3@. We use variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.

@@

Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and continues with the instruction at @A0@.
This @drop@ instruction is used to coordinate processes when performing fusion and concurrent evaluation; the next element of a stream may only be pulled after all consuming processes have pulled and dropped the current element.
% This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm and concurrent evaluation, but has no direct operational meaning.
% We discuss @drop@ further in \TODO{ref}.

@@

As this process merges infinite streams, if we execute it using a finite prefix then the result state will be an intermediate one that may not yet have pushed all available output. For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output. The process will arrive at instruction @E2@ which blocks waiting for the next value to be pulled from @s2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}

@@

% -----------------------------------------------------------------------------
\clearpage{}
\subsection{Fusion}
Our fusion algorithm takes two process state machines and produces a new one that performs the job of both. For example, suppose we need to produce a single machine that performs the job of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}. We want a result machine that performs the job of a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@. Relative to the @uniquesUnion@ example we will associate a stream named @s1@ with the first array (@arr1@) and a stream named @s2@ with the second.

@@

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine. In Fig.\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1' is set to `have' in @F3@ after we have set `x1 = b1'.

@@

% -----------------------------------------------------------------------------
\subsection{Fused Result}

Fig.~\ref{fig:Process:Fused} shows the result of fusing @group@ and @merge@ together, after some manual simplification.
The redundant @jump@s at @F1@ and @F2@ have been removed, and the variables @v@ and @x1@ have been merged into a single variable, @x1@.

\amos{Or is it better to have the actual unsimplified result of fusion? It will be bigger, but don't have to handwave.}

\begin{figure}
\begin{code}
            process
            { ins:    { s1, s2 }
            , outs:   { s3, s4 }
            , heap:   { f = T, l = 0, x1 = 0, x2 = 0}
            , label:  F0
            , instrs: { F0 = pull s1 x1              F3 {}
                      , F3 = case (f || (l /= x1))   F4 {} F5 {}
                      , F4 = push s3 x1              F5 { l = x1, f = F }
                      , F5 = pull s2 x2              F6 {}

                      , F6 = case (x1 < x2)          A0 {} B0 {}

                      , A0 = push s4 x1              A1 {}
                      , A1 = drop s1                 A2 {}
                      , A2 = pull s1 x1              A3 {}
                      , A3 = case (f || (l /= x1))   A4 {} F6 {}
                      , A4 = push s3 x1              F6 { l = x1, f = F }

                      , B0 = push s4 x2              B1 {}
                      , B1 = drop s2                 B2 {}
                      , B2 = pull s2 x2              F6 {} } }
\end{code}
\caption{Fusion of group and merge}
\label{fig:Process:Fused}
\end{figure}

@@

A process defines a stream computation, taking any number of input streams and producing at least one output stream.
The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
The input states are explained in detail later in \S\ref{s:Process:Eval}.

The output streams are in some sense ``owned'' by the process that produces them: while a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
A process may, however, produce multiple output streams.

The heap is used for evaluation of expressions.
Each process has its own private heap, therefore the only communication between processes occurs by streams.

The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
Instructions can pull from a stream, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.

@@

All instructions take as an argument the next label to jump to, as well as any variable updates that should be performed on the private heap at the same time.
% Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.

A process network is a set of multiple processes that can be evaluated concurrently.
Any inputs that are not produced as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

@@

We now describe evaluation of processes and process networks.
We split evaluation into three main parts:
\begin{itemize}
\item Injection, where values are inserted into a process' input buffer.
Injection is only possible when the process input buffer is empty.
\item Shaking, where a process takes a step from one label to another.
Shaking a process results in an updated process as well as an output message.
If a process pushes to a stream, the push value must be able to be injected to other processes.
\item Feeding, where an environment of input values are fed to the processes, and output values are collected.
This is the `top level' of evaluation that uses both injection and shaking.
\end{itemize}
